Supporting Information: Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-06-01 Ben_Manuscripts/hdphmm/Supporting_Information.tex
Incorporate Chris' suggestions
comments on supporting.

  The state sequence appears to converge within 500--1000 IHMM iterations. To
  be sure, we ran 2000 iterations of the IHMM algorithm in order to arrive at a 
  finalized state sequence. Stabilization of the VAR parameters indicates 
  that the state sequence has converged. In Figure~\ref{fig:convergence3d}, we plot
  the diagonal entries of the $A$ and $\Sigma$ matrices as well as the entries of
  the mean vector $\mathbf{c}$, for three different states identified in a 3D 
  methanol trajectory. States with a relatively small number of observations are
  more highly influenced by the boundaries of the state segments and generally 
  lead to much higher variance of the $A$ and $\Sigma$ parameters. In these cases,
  the mean is a more reliable indicator of convergence since stabilized means 
  imply that the fluctuations assigned to each state come from motion about a 
  static location.

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{overestimate_ACH_21.pdf}
  \caption{}\label{fig:msd_overestimate}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_before_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_before}
  \end{subfigure}
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{msd_improvement_ACH_21.pdf}
  \caption{}\label{fig:msd_improvement}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_after_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_after}
  \end{subfigure}
  \caption{(a, b) In some cases, the state segmentation leads to over- or 
  under-prediction of the MSD, usually due to poor initialization of the state
  sequence. In (b), state transitions are too frequent, particularly between 
  the light blue and light green states, and ultimately causes an over-prediction
  of the MD MSD. (c) We can greatly improve MSD predictions by giving the HDP-AR-HMM a good
  guess at the initial state sequence. (d) A better initial state segmentation
  via seeding, gives more reasonable estimates of the final state sequence
  as illustrated in Figure~\ref{fig:seed_sequence}.}\label{fig:improvement_state_sequence}
  \end{figure}

@@

  We can help to further narrow our decision of linkage criteria by analyzing 
  the parameters of the clusters which it produces. In particular, we should 
  look for well-distinguished clusters on the radial means so that it will be
  easier to connect solute behavior to the membrane pore structure. One way to
  concisely visualize the radial clusters is by their spread. An optimal number
  of clusters should have the same spread in radial means as the unclustered
  data. In Figure~\ref{fig:rspread}, we show that `ward' linkage results in 
  the highest spread in the radials means of the clusters across a range of
  4--30 total clusters. For a low to intermediate number of clusters, the 
  spread of `ward' clusters is significantly larger than of the other linkage
  methods. Therefore, we have chosen to use `ward' clustering.

  We use the Silhouette score in order to aid us in choosing the total number of
  clusters. In Figure~\ref{fig:silhouette_MET}, we plot the Silhouette score as
  a function of the total number of clusters using the `ward' linkage criteria. 
  A Silhouette score of 1 indicates the best clustering possible, values near
  0 indicate overlapping clusters, and negative values indicate that samples belong to
  the wrong clusters. For the methanol parameters, the Silhouette score drops off
  quickly with number of clusters before plateauing around 7 total clusters. 
  However, given the low spread in $\mu_r$, it is not clear that using 7 or less 
  clusters will give useful clustering results.

  \begin{figure}
  \centering
  \begin{subfigure}{0.53\textwidth}
  \includegraphics[width=\textwidth]{rspread_nclusters.pdf}
  \caption{}\label{fig:rspread}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_MET.pdf}
  \caption{}\label{fig:silhouette_MET}
  \end{subfigure}
  \caption{(a) The `ward' linkage criteria maintains the highest standard 
  deviation in radial means, $\mu_r$, of the clusters for a given total 
  number of clusters as compared to other linkage criteria. (b) The 
  Silhouette score, used to evaluate the quality of clustering with the 
  `ward' distance criteria drops precipitously as a function of the chosen
  total number of clusters before plateauing.}\label{fig:idk}
  \end{figure}

  We incorporate qualitative feedback into the decision on the number of 
  clusters. Despite its high Silhouette score, Figure~\ref{fig:4cluster_state_sequence}
  illustrates that using 4 clusters clearly does not distinguish dynamical 
  modes. We do not see visually acceptable clustering on the example trajectory
  until we use at least twenty total clusters (Figure~\ref{fig:20cluster_state_sequence}).

  \begin{figure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_4.pdf}
  \caption{4 total clusters}\label{fig:4cluster_state_sequence}
  \end{subfigure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_10_nolabel.pdf}
  \caption{10 total clusters}\label{fig:20cluster_state_sequence}
  \end{subfigure}
  \caption{Although the Silhouette score for four total clusters is high, we do
  not see adequate distinction between clusters until we group the parameters
  from the 24 trajectories into at least 10 total clusters.}\label{fig:clustered_state_sequences}
  \end{figure}

@@

  Realizations of the HDP-AR-HMM underestimate the MD MSD of urea due to MD
  trajectories with axial motion correlation that is likely uncharacteristic and
  unable to be captured by ensembles of stochastic HDP-AR-HMM realizations. One
  such trajectory is plotted in Figure~\ref{fig:urea_underestimate}. Axial hops 
  appear positively correlated, jumping consistently in the negative $z$ direction, 
  leading to a high MSD. The parabolic trajectory of the MD MSD curve implies
  underlying superdiffusion which would require some kind of facilitating
  driving force. It is likely that running the trajectory longer would reduce
  the slope of the MSD curve, making it more consistent with the HDP-AR-HMM 
  prediction.

@@

  At extremely short time lags, the MSD is generally over-predicted by the HDP-AR-HMM
  (see Figure~\ref{fig:unclustered_msd_shortlag}) because the VAR(1) model assumes
  multivariate Gaussian noise while the MD data suggests the noise is better 
  modeled by more general L\'evy stable noise. The first time lag of the MSD is
  equivalent to the variance of all observed fluctuations. 
  The HDP-AR-HMM simulation appears to over-estimate the variance because the actual 
  distribution of solute fluctuations has heavy tails with a smaller variance (Figure~\ref{fig:emission_widths}).
  The total distribution of hops is expected to be heavy tailed because it is a
  combination of hop distributions from many states with different variances. More
  interestingly, in Figure ~\ref{fig:state_emission_widths}, we show that the 
  emission distribution of a single state picked out by the HDP-AR-HMM is fit better by
  a more general L\'evy distribution while the emissions from HDP-AR-HMM realizations are
  Gaussian, as expected. An improvement to the HDP-AR-HMM that may allow dynamics more 
  faithful to MD would use an autoregressive model with L\'evy stable noise. However,
  this would be less computationally efficient due to lack of conjugate priors.

@@

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{local_densities.pdf}
  \caption{We measured the local number density of heavy atoms surrounding solutes in each
  state. In blue, we plot the local number density experienced by solutes averaged over all
  frames in which the solute occupied the given state. We broke down this overall average 
  number density into number densities measured from instances when the state is occupied 
  for long (orange) versus short(red) dwell time states. We define define `long' dwells times
  as the 50th percentile of all dwell times in the given state, and short as all others. It
  is clear that solutes which experience longer dwells times are also surrounded by more 
  heavy atoms on average.}\label{fig:local_densities}
  \end{figure}

\end{document}

>>>>

  The state sequence appears to converge within 500--1000 IHMM iterations. To
  be certain, we ran 2000 iterations of the IHMM algorithm in order to arrive at a 
  finalized state sequence. Stabilization of the VAR parameters indicates 
  that the state sequence has converged. In Figure~\ref{fig:convergence3d}, we plot
  the diagonal entries of the $A$ and $\Sigma$ matrices as well as the entries of
  the mean vector $\mathbf{c}$, for three different states identified in a 3D 
  methanol trajectory. States with a relatively small number of observations are
  more highly influenced by the boundaries of the state segments and generally 
  lead to much higher variance of the $A$ and $\Sigma$ parameters. In these cases,
  the mean is a more reliable indicator of convergence since stabilized means 
  imply that the fluctuations assigned to each state come from motion about a 
  static location.

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{overestimate_ACH_21.pdf}
  \caption{}\label{fig:msd_overestimate}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_before_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_before}
  \end{subfigure}
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{msd_improvement_ACH_21.pdf}
  \caption{}\label{fig:msd_improvement}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_after_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_after}
  \end{subfigure}
  \caption{(a, b) In some cases, the state segmentation leads to over- or 
  under-prediction of the MSD, usually due to poor initialization of the state
  sequence. In (b), state transitions are too frequent, particularly between 
  the light blue and light green states, and ultimately causes an over-prediction
  of the MD MSD. (c) We can greatly improve MSD predictions by giving the HDP-AR-HMM a good
  guess at the initial state sequence. (d) A better initial state segmentation
  via seeding gives more reasonable estimates of the final state sequence
  as illustrated in Figure~\ref{fig:seed_sequence}.}\label{fig:improvement_state_sequence}
  \end{figure}

@@

  We can help to further narrow our decision of linkage criteria by analyzing 
  the parameters of the clusters which it produces. In particular, we should 
  look for well-distinguished clusters on the radial means so that it will be
  easier to connect solute behavior to the membrane pore structure. One way to
  concisely visualize the radial clusters is by their spread. An optimal number
  of clusters should have the same spread in radial means as the unclustered
  data. In Figure~\ref{fig:rspread}, we show that `ward' linkage results in 
  the highest spread in the radial means of the clusters across a range of
  4--30 total clusters. For a low to intermediate number of clusters, the 
  spread of `ward' clusters is significantly larger than of the other linkage
  methods. Therefore, we have chosen to use `ward' clustering.

  We use the silhouette score in order to aid us in choosing the total number of
  clusters. In Figure~\ref{fig:silhouette_MET}, we plot the silhouette score as
  a function of the total number of clusters using the `ward' linkage criteria. 
  A wilhouette score of 1 indicates the best clustering possible, values near
  0 indicate overlapping clusters, and negative values indicate that samples belong to
  the wrong clusters. For the methanol parameters, the silhouette score drops off
  quickly with number of clusters before plateauing around 7 total clusters. 
  However, given the low spread in $\mu_r$, it is not clear that using 7 or less 
  clusters will give useful clustering results.

  \begin{figure}
  \centering
  \begin{subfigure}{0.53\textwidth}
  \includegraphics[width=\textwidth]{rspread_nclusters.pdf}
  \caption{}\label{fig:rspread}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_MET.pdf}
  \caption{}\label{fig:silhouette_MET}
  \end{subfigure}
  \caption{(a) The `ward' linkage criteria maintains the highest standard 
  deviation in radial means, $\mu_r$, of the clusters for a given total 
  number of clusters as compared to other linkage criteria. (b) The 
  silhouette score, used to evaluate the quality of clustering with the 
  `ward' distance criteria drops precipitously as a function of the chosen
  total number of clusters before plateauing.}\label{fig:idk}
  \end{figure}

  We incorporate qualitative feedback into the decision on the number of 
  clusters. Despite its high silhouette score, Figure~\ref{fig:4cluster_state_sequence}
  illustrates that using 4 clusters clearly does not distinguish dynamical 
  modes. We do not see visually acceptable clustering on the example trajectory
  until we use at least twenty total clusters (Figure~\ref{fig:20cluster_state_sequence}).

  \begin{figure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_4.pdf}
  \caption{4 total clusters}\label{fig:4cluster_state_sequence}
  \end{subfigure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_10_nolabel.pdf}
  \caption{10 total clusters}\label{fig:20cluster_state_sequence}
  \end{subfigure}
  \caption{Although the silhouette score for four total clusters is high, we do
  not see adequate distinction between clusters until we group the parameters
  from the 24 trajectories into at least 10 total clusters.}\label{fig:clustered_state_sequences}
  \end{figure}

@@

  Realizations of the HDP-AR-HMM underestimate the MD MSD of urea due to MD
  trajectories with axial motion correlation that is likely uncharacteristic and
  unable to be captured by ensembles of stochastic HDP-AR-HMM realizations. One
  such trajectory is plotted in Figure~\ref{fig:urea_underestimate}. Axial hops 
  appear positively correlated, jumping consistently in the negative $z$ direction, 
  leading to a high MSD. The parabolic trajectory of the MD MSD curve implies
  underlying super-diffusion which would require some kind of facilitating
  driving force. It is likely that running the trajectory longer would reduce
  the slope of the MSD curve, making it more consistent with the HDP-AR-HMM 
  prediction.

@@

  At extremely short time lags, the MSD is generally over-predicted by the HDP-AR-HMM
  (see Figure~\ref{fig:unclustered_msd_shortlag}) because the VAR(1) model assumes
  multivariate Gaussian noise while the MD data suggests the noise is better 
  modeled by more general L\'evy stable noise. The first time lag of the MSD is
  equivalent to the variance of all observed fluctuations. 
  The HDP-AR-HMM simulation appears to over-estimate the variance because the actual 
  distribution of solute fluctuations has 
  %MRS: heavy tails should mean larger variance? Clarify?
  heavy tails with a smaller variance (Figure~\ref{fig:emission_widths}).
  The total distribution of hops is expected to be heavy tailed because it is a
  combination of hop distributions from many states with different variances. More
  interestingly, in Figure ~\ref{fig:state_emission_widths}, we show that the 
  emission distribution of a single state picked out by the HDP-AR-HMM is fit better by
  a more general L\'evy distribution while the emissions from HDP-AR-HMM realizations are
  Gaussian, as expected. An improvement to the HDP-AR-HMM that may allow dynamics more 
  faithful to MD would use an autoregressive model with L\'evy stable noise. However,
  this would be less computationally feasible due to lack of conjugate priors.

@@

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{local_densities.pdf}
  \caption{We measured the local number density of heavy atoms surrounding solutes in each
  state. In blue, we plot the local number density experienced by solutes averaged over all
  frames in which the solute occupied the given state. We broke down this overall average 
  number density into number densities measured from instances when the state is occupied 
  for long (orange) versus short (red) dwell time states. We define define `long' dwells times
  as the 50th percentile of all dwell times in the given state, and short as all others. It
  is clear that solutes which experience longer dwells times are also surrounded by more 
  heavy atoms on average.}\label{fig:local_densities}
  \end{figure}

  \subsection{Predominant States}\label{section:state_prevelance}

  \begin{figure}[h]
  \centering
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{prevalence_GCL.pdf}
  \caption{ethylene glycol}\label{fig:prevalence_GCL}
  \end{subfigure}
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{prevalence_URE.pdf}
  \caption{urea}\label{fig:prevalence_URE}
  \end{subfigure}
  \begin{subfigure}{0.325\textwidth}
  \includegraphics[width=\textwidth]{prevalence_ACH.pdf}
  \caption{acetic acid}\label{fig:prevalence_ACH}
  \end{subfigure}
  \caption{MD trajectories sample the clustered states to varying degrees and
  for varied amounts of time. For each solute in (a), (b) and (c), we plot
  the total number of trajectories that sample a given state (blue) as well
  as the fraction of the total simulation time spent in that state.}
  \label{fig:state_prevalence}
  \end{figure}

\end{document}
% LocalWords:  IHMM pdf prior's solute MSD AR HMM ACH Dirichlet delocalize hmm
% LocalWords:  Gaussians methanol's scikit plateauing nolabel MSDs urea's URE
% LocalWords:  shortlag evy BJC SI msds msd nclusters Solutes solutes alkane
